---
title: "ST 558 Project 2"
author: "Hannah Park"
date: "10/15/2020"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(GGally)
```

# Introduction  

You should have an introduction section that briefly describes the data and the variables you have to work with (no need to discuss all of them, just the ones you want to use). **If you are analyzing the bike share data, do not use the casual and registered variables to do any modeling!**  

You should also mention the purpose of your analysis and the methods you’ll use (no need to detail them here) for analysis.


# Data  
```{r}
# Read in data
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00275/Bike-Sharing-Dataset.zip"
download.file(url, "Bike-Sharing-Dataset.zip")

unzip("Bike-Sharing-Dataset.zip", exdir = "./Data")
df.bike <- read_csv("/Data/day.csv")

df.bike.day <- df.bike %>%
  select(-casual, -registered) %>%
  filter(weekday == 1)

# Randomply sample from the data 
# Form training and test sets
train <- sample(1:nrow(df.bike.day), size = nrow(df.bike.day)*0.7)
test <- dplyr::setdiff(1:nrow(df.bike.day), train)
df.train <- df.bike.day[train, ]
df.test <- df.bike.day[test, ]
```

# Summarizations  
```{r}
ggplot(df.train, aes(x = dteday, y = cnt)) +
  geom_point(aes(colour = factor(season)))

df.train %>%
  gather(temp, atemp, hum, windspeed, key = "var", value = "value") %>%
  ggplot(aes(x = value, y = cnt, color = factor(yr), shape = factor(season))) +
  geom_point() +
  geom_smooth(aes(group = factor(yr))) +
  facet_wrap(~var, scales = "free") 
```

