---
title: "ST 558 Project 2"
author: "Hannah Park"
date: "10/15/2020"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(tidyverse)
library(ggplot2)
library(GGally)
library(corrplot)
```

# Introduction  

You should have an introduction section that briefly describes the data and the variables you have to work with (no need to discuss all of them, just the ones you want to use). **If you are analyzing the bike share data, do not use the casual and registered variables to do any modeling!**  

You should also mention the purpose of your analysis and the methods you’ll use (no need to detail them here) for analysis.

See Figure \ref{fig:correlation-plot}.

# Data  
```{r, message = FALSE, warning = FALSE}
# Read in data
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00275/Bike-Sharing-Dataset.zip"
download.file(url, "Bike-Sharing-Dataset.zip")

unzip("Bike-Sharing-Dataset.zip", exdir = "./Data")
df.bike <- read_csv("/Data/day.csv") %>%
  select(-instant, -casual, -registered) %>%
  mutate(dayofweek = recode(weekday,
                            `0` = "Sunday",
                            `1` = "Monday",
                            `2` = "Tuesday",
                            `3` = "Wednesday",
                            `4` = "Thursday",
                            `5` = "Friday",
                            `6` = "Saturday"))
  #Combine variables temp and atemp

df.bike.day <- df.bike %>%
  filter(weekday == 1)

#Randomply sample from the data 
#Form training and test sets
train <- sample(1:nrow(df.bike.day), size = nrow(df.bike.day)*0.7)
test <- dplyr::setdiff(1:nrow(df.bike.day), train)
df.train <- df.bike.day[train, ]
df.test <- df.bike.day[test, ]
```

# Summarizations  
## Full Data  

### Correlation Plot  
```{r correlation-plot, fig.cap = "Correlation Plot"}
corrplot(cor(df.bike[,2:13]))
```

```{r}

#Qualitative variables: Contingency tables


#Quantitative variables: Summary statistics




#Histograms of quantitative variables
df.bike %>%
  gather(temp, atemp, hum, windspeed, key = "var", value = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram() +
  facet_wrap(~var, scales = "free") 

#Scatterplot of response variable(cnt) over days
ggplot(df.bike, aes(x = dteday, y = cnt)) +
  geom_point(aes(colour = factor(holiday)))

```

## Specific Day of the Week Data  
```{r}

df.train %>%
  gather(temp, atemp, hum, windspeed, key = "var", value = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram() +
  facet_wrap(~var, scales = "free") 



df.train %>%
  gather(temp, atemp, hum, windspeed, key = "var", value = "value") %>%
  ggplot(aes(x = value, y = cnt, color = factor(yr), shape = factor(season))) +
  geom_point() +
  geom_smooth(aes(group = factor(yr))) +
  facet_wrap(~var, scales = "free") 
```

